---
title: "Project1: Interim Data Analysis Plan"
author: "Eduardo Romo"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
    toc: false
fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{float}
  - \usepackage{graphicx}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \setlength{\parindent}{2em}
  - \setlength{\parskip}{0pt}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6.5, fig.height = 4.5
)

library(knitr)
library(kableExtra)
library(stringr)
library(tidyverse)
hiv_data <- read_csv("../Data/hiv_6624_final.csv") %>%
  select(-`...1`)

#see unique race and education levels in data to see which labels are needed
unique(hiv_data$RACE)
unique(hiv_data$EDUCBAS)

#collapse race variable to White vs Other
hiv_data <- hiv_data %>%
  mutate(
    race_cat = factor(case_when(
      RACE %in% c(1, 2) ~ 1,
      RACE %in% c(3, 4, 7, 8) ~ 2
    ),
  levels = c(1, 2), labels = c("White", "Other")))
#collapse education variable to College Degree+ vs Non College Degree
hiv_data <- hiv_data %>%
  mutate(
    edu_cat = factor(case_when(
      EDUCBAS %in% c(1, 2, 3, 4) ~ 1,
      EDUCBAS %in% c(5, 6, 7) ~ 2
    ),
    levels = c(1, 2), labels = c("No College Degree", "College Degree")
    )
  )
#recode abnormal BMI values so that they are missing/NAs
hiv_data <- hiv_data %>%
  mutate(
    BMI = if_else(BMI <= 0 | BMI > 250, NA_real_, BMI)
  )
#create labelled versions of categorical variables
hiv_data <- hiv_data %>%
  mutate(
    smoke_cat = factor(
      SMOKE,
      levels = c(1, 2, 3),
      labels = c("Never smoked",
                 "Former smoker",
                 "Current smoker")
    ),
    adh_cat = factor(
      ADH,
      levels = c(1, 2, 3, 4),
      labels = c("100\\%",
                 "95-99\\%",
                 "75-94\\%",
                 "<75\\%")
    ),
    hard_drugs_cat = factor(
      hard_drugs,
      levels = c(0, 1),
      labels = c("No",
                 "Yes")
    )
  )
#create analysis dataset
hivp1_data <- hiv_data %>%
  #keep only baseline and year 2 measurements
  filter(years %in% c(0,2)) %>%
  #keep only the columns we need
  #covariates will become *_0 and *_2 after pivoting
  pivot_wider(
    id_cols = newid,
    names_from = years,
    values_from = c(AGG_MENT, AGG_PHYS, LEU3N, VLOAD, income, BMI, SMOKE, smoke_cat,
                    ADH, adh_cat, RACE, race_cat, EDUCBAS, edu_cat, age, hard_drugs,
                    hard_drugs_cat),
    names_sep = "_"
  ) %>%
  #compute year2 - baseline diffs for outcomes
  mutate(
    d_AGG_MENT = AGG_MENT_2 - AGG_MENT_0,
    d_AGG_PHYS = AGG_PHYS_2 - AGG_PHYS_0,
    d_LEU3N = LEU3N_2 - LEU3N_0,
    d_VLOAD = VLOAD_2 - VLOAD_0
  ) %>%
  select(newid, ends_with("_0"), ends_with("_2"), starts_with("d_"))

# try log10 transformation of ratio of VLOAD_2 and VLOAD_0
hivp1_data <- hivp1_data %>%
  mutate(
    r_logVLOAD = log10(VLOAD_2/VLOAD_0),
    logVLOAD_0 = log10(VLOAD_0)
  )
#exclude observations that have missing values for outcomes for the descriptive
#summaries and analysis
hivp1_data <- hivp1_data %>%
  drop_na(d_AGG_MENT, d_AGG_PHYS, d_LEU3N, r_logVLOAD)

### ---- code for creating table 1 --------
# --- helper functions ---
fmt_n_pct <- function(n, denom) {
  if (is.na(n) | is.na(denom) | denom == 0) return(NA_character_)
  sprintf("%d (%.1f\\%%)", n, 100 * n / denom)
}

fmt_mean_sd <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  sprintf("%.1f (%.1f)", mean(x), sd(x))
}

fmt_missing <- function(x) {
  n_miss <- sum(is.na(x))
  p_miss <- mean(is.na(x)) * 100
  sprintf("%d (%.1f\\%%)", n_miss, p_miss)
}

# --- main Table 1 function ---
make_table1 <- function(df,
                        group_var = hard_drugs_cat,
                        cont_vars = c("BMI_0","age_0", "LEU3N_0", "logVLOAD_0", "AGG_MENT_0", "AGG_PHYS_0"),
                        cat_vars  = c("smoke_cat_0","adh_cat_2","race_cat_0","edu_cat_0"),
                        var_labels = list(
                          BMI_0    = "BMI",
                          age_0    = "Age",
                          smoke_cat_0 = "Smoking status",
                          adh_cat_2   = "Adherence category (year 2)",
                          race_cat_0  = "Race/ethnicity",
                          edu_cat_0   = "Education",
                          LEU3N_0 = "CD4+ Count",
                          logVLOAD_0 = "Log Viral Load",
                          AGG_MENT_0 = "Mental Quality of Life",
                          AGG_PHYS_0 = "Physical Quality of Life"
                        )) {
  
  # ensure group is a factor with nice labels
  df <- df %>%
    mutate(.group = as.factor({{ group_var }}))
  
  # group counts (non-missing group)
  n_overall <- nrow(df)
  n_bygrp <- df %>%
    filter(!is.na(.group)) %>%
    count(.group, name = "N")
  
  # ---------- continuous summaries ----------
  cont_tbl <- lapply(cont_vars, function(v) {
    lab <- var_labels[[v]] %||% v
    
    overall_mean_sd   <- fmt_mean_sd(df[[v]])
    overall_miss      <- fmt_missing(df[[v]])
    
    bygrp <- df %>%
      group_by(.group) %>%
      summarise(
        mean_sd = fmt_mean_sd(.data[[v]]),
        miss    = fmt_missing(.data[[v]]),
        .groups = "drop"
      ) %>%
      right_join(n_bygrp, by = ".group") %>%
      arrange(.group)
    
    # build rows: Mean (SD), Missing
    out <- tibble(
      Variable = c(lab, ""),
      Level = c("Mean (SD)", "Missing"),
      Overall = c(overall_mean_sd, overall_miss)
    )
    
    # add each group as columns (No/Yes)
    for (g in levels(df$.group)) {
      val_mean_sd <- bygrp$mean_sd[bygrp$.group == g]
      val_miss    <- bygrp$miss[bygrp$.group == g]
      out[[as.character(g)]] <- c(val_mean_sd, val_miss)
    }
    
    out
  }) %>% bind_rows()
  
  # ---------- categorical summaries ----------
  cat_tbl <- lapply(cat_vars, function(v) {
    lab <- var_labels[[v]] %||% v
    
    # denominators for percents: non-missing within each column
    denom_overall <- sum(!is.na(df[[v]]))
    denom_bygrp <- df %>%
      group_by(.group) %>%
      summarise(denom = sum(!is.na(.data[[v]])), .groups = "drop")
    
    # all levels (excluding NA)
    levs <- df[[v]] %>% as.factor() %>% levels()
    
    # counts overall
    overall_counts <- df %>%
      filter(!is.na(.data[[v]])) %>%
      count(.data[[v]], name = "n") %>%
      rename(Level = 1)
    
    # counts by group
    grp_counts <- df %>%
      filter(!is.na(.group)) %>%
      count(.group, .data[[v]], name = "n") %>%
      rename(Level = 2)
    
    # base rows for each level
    out <- tibble(
      Variable = c(lab, rep("", length(levs) - 1), ""),
      Level = c(levs, "Missing")
    )
    
    # Overall column
    out$Overall <- c(
      sapply(levs, function(L) {
        nL <- overall_counts$n[overall_counts$Level == L]
        nL <- ifelse(length(nL) == 0, 0, nL)
        fmt_n_pct(nL, denom_overall)
      }),
      fmt_missing(df[[v]])
    )
    
    # group columns
    for (g in levels(df$.group)) {
      denom_g <- denom_bygrp$denom[denom_bygrp$.group == g]
      denom_g <- ifelse(length(denom_g) == 0, NA_integer_, denom_g)
      
      out[[as.character(g)]] <- c(
        sapply(levs, function(L) {
          nL <- grp_counts$n[grp_counts$.group == g & grp_counts$Level == L]
          nL <- ifelse(length(nL) == 0, 0, nL)
          fmt_n_pct(nL, denom_g)
        }),
        fmt_missing(df[[v]][df$.group == g])
      )
    }
    
    out
  }) %>% bind_rows()
  
  # ---------- sample size row ----------
  Ns <- tibble(
    Variable = "Sample size",
    Level = "N",
    Overall = as.character(n_overall)
  )
  for (g in levels(df$.group)) {
    Ns[[as.character(g)]] <- as.character(sum(df$.group == g, na.rm = TRUE))
  }
  
  # combine
  bind_rows(Ns, cont_tbl, cat_tbl)
}

tab1 <- make_table1(
  hivp1_data,
  group_var = hard_drugs_cat_0,   # baseline hard drug use category column after pivot
  cont_vars = c("BMI_0","age_0", "LEU3N_0", "logVLOAD_0", "AGG_MENT_0", "AGG_PHYS_0"),
  cat_vars  = c("smoke_cat_0","adh_cat_2","race_cat_0","edu_cat_0")
)

# Nice column names
colnames(tab1) <- str_replace_all(colnames(tab1), "^No$", "No hard drugs")
colnames(tab1) <- str_replace_all(colnames(tab1), "^Yes$", "Yes hard drugs")

#code that will be used to knit table1 to pdf from rmarkdown in chunk
table1 <- kbl(
  tab1,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  align = "llccc",
  caption = "Baseline characteristics stratified by baseline hard drug use."
  ,
  escape = FALSE,
  linesep = "") %>%
  kable_styling(latex_options = c("hold_position", "repeat_header"), font_size = 8) %>%
  column_spec(1, width = "4.5cm") %>%
  column_spec(2, width = "2.5cm") %>%
  column_spec(3:ncol(tab1), width = "2.2cm")
```

# Introduction
Highly Active Antiretroviral Therapy (HAART) is the standard treatment regimen for patients diagnosed with HIV. HAART suppresses viral replication, improving immune function as reflected by increases in CD4+ T cell counts, and substantially reducing mortality and disease progression. The Multicenter AIDS Cohort Study is an ongoing prospective cohort study of the natural and treated histories of HIV-1 infection in homosexual and bisexual men in four major cities in the United States. The dataset used in this project includes annual longitudinal laboratory and quality of life measurements, along with demographic and clinical health information, collected for up to eight years after HIV-infected men initiated HAART. 

Investigators are interested in conducting a secondary data analysis to evaluate whether hard drug use at baseline is associated with poorer treatment response following HAART initiation. The primary scientific hypothesis is that subjects who report hard drug use at baseline will have worse treatment response two years after initiating HAART compared to subjects who do not report hard drug use. Treatment response will be evaluated using four complementary outcomes: viral load, CD4+ T cell count, and physical and mental quality of life scores.

The corresponding statistical hypotheses evaluate whether mean treatment response at two years differs by baseline hard drug use status for each outcome, adjusting for age, BMI, smoking status, education, and race/ethnicity. For each outcome, the null hypothesis is that there is no difference in mean treatment response between subjects who report hard drug use at baseline and those who do not, while the alternative hypothesis is that a difference exists. These hypotheses will be evaluated using both Bayesian and frequentist methods.

# Preliminary Methods
The dataset includes 715 unique patients. The primary exposure variable (baseline hard drug use) and several adjustment variables (race/ethnicity, education, smoking status, and treatment adherence) were recorded numerically in the raw dataset but represent categorical variables. These variables were converted to factor variables with descriptive labels to improve interpretability in descriptive summaries and modeling.

Because the primary research question focuses on treatment response between baseline (year 0) and two years after HAART initiation (year 2), the analysis dataset was created by restricting observations to years 0 and 2 and restructuring the data to create patient-level outcomes. Change-from-baseline outcomes were constructed for CD4+ T cell count and physical and mental quality of life scores as the difference between year 2 and year 0 values. Due to skewness in viral load change, the natural logarithm of the ratio of viral load at year 2 to baseline viral load was used instead. These outcomes allow estimation of average treatment response differences between exposure groups using regression-based methods.

Primary analyses will evaluate the association between baseline hard drug use and treatment response using multivariable linear regression models for each outcome, adjusting for age, BMI, smoking status, education, and race/ethnicity. Both frequentist and Bayesian linear regression models will be fit for each outcome to allow comparison of inference across frameworks. Frequentist results will be summarized using estimated coefficients, standard errors, 95% confidence intervals, and p-values. Bayesian results will be summarized using posterior means, 95% credible intervals, and posterior probabilities that regression coefficients differ from zero.

Additionally, mediation analyses will be conducted using linear regression-based approaches to evaluate whether treatment adherence mediates the relationship between baseline hard drug use and treatment response. Both Bayesian and frequentist mediation models will be fit and compared using corresponding interval estimates and measures of statistical or posterior evidence.

# Preliminary Analysis

\newpage

```{r, echo=FALSE, cache=FALSE}
table1
```

